### 2021-07-27 tex/kits.tex
[ refactor ] lagda to separate file
[ redraft ] kits introduction

\def\prefix{../agda/latex}
\CatchFileBetweenTags{\Var}{\prefix/SimpleKits.tex}{Var}
\CatchFileBetweenTags{\Term}{\prefix/SimpleKits.tex}{Term}
\CatchFileBetweenTags{\Ren}{\prefix/SimpleKits.tex}{Ren}
\CatchFileBetweenTags{\bindRen}{\prefix/SimpleKits.tex}{bindRen}
\CatchFileBetweenTags{\rename}{\prefix/SimpleKits.tex}{rename}
\CatchFileBetweenTags{\leftTerm}{\prefix/SimpleKits.tex}{leftTerm}
\CatchFileBetweenTags{\Sub}{\prefix/SimpleKits.tex}{Sub}
\CatchFileBetweenTags{\bindSub}{\prefix/SimpleKits.tex}{bindSub}
\CatchFileBetweenTags{\substitute}{\prefix/SimpleKits.tex}{substitute}
\CatchFileBetweenTags{\Env}{\prefix/SimpleKits.tex}{Env}
\CatchFileBetweenTags{\RenSub}{\prefix/SimpleKits.tex}{RenSub}
\CatchFileBetweenTags{\Kit}{\prefix/SimpleKits.tex}{Kit}
\CatchFileBetweenTags{\trav}{\prefix/SimpleKits.tex}{trav}
\CatchFileBetweenTags{\renKit}{\prefix/SimpleKits.tex}{renKit}
\CatchFileBetweenTags{\subKit}{\prefix/SimpleKits.tex}{subKit}

\CatchFileBetweenTags{\RenGD}{\prefix/SimpleKits.tex}{RenGD}
\CatchFileBetweenTags{\RenGADA}{\prefix/SimpleKits.tex}{RenGADA}
\CatchFileBetweenTags{\GTh}{\prefix/SimpleKits.tex}{GTh}
\CatchFileBetweenTags{\DTh}{\prefix/SimpleKits.tex}{DTh}
\CatchFileBetweenTags{\GD}{\prefix/SimpleKits.tex}{GD}

@@

\begin{mathpar}
  \inferrule*[right=Subst]
  {%
    \inferrule*[right=$\to$-E]
    {%
      \inferrule*[right=Var]{~}{A \to B, A \vdash A \to B}
      \\
      \inferrule*[Right=Var]{~}{A \to B, A \vdash A}
    }
    {A \to B, A \vdash B}
    \\
    B \vdash C
  }
  {A \to B, A \vdash C}
\end{mathpar}

\begin{displaymath}
  \begin{prooftree}
    \infer0[Var]{A \to B, A \vdash A \to B}
    \infer0[Var]{A \to B, A \vdash A}
    \infer2[$\to$-E]{A \to B, A \vdash B}
    \hypo{B \vdash C}
    \infer2[Subst]{A \to B, A \vdash C}
  \end{prooftree}
\end{displaymath}

\subsection{Proofs of renaming and substitution}

We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
terms.
Beside base types, the only type former we have is the function type constructor
\AgdaInductiveConstructor{\_`→\_}.
Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
(\AgdaDatatype{Ty}).
Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
embeds types into contexts.
Typed variables in a context are given by \AgdaRecord{\_∋\_}.
A variable in
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.

\Var{}
\Term{}

@@

In the following, \AgdaFunction{ren} gives the action of a renaming on terms.
We replace variables by new variables given by the renaming \AgdaBound{$\rho$}.
For applications, we rename each subterm using the same renaming.
For $\lambda$-abstractions, we want to rename the body, but the renaming we
have has type \RenGD{}, whereas the renaming we need is of type \RenGADA{}.
To make up the difference, we introduce the \AgdaFunction{bindRen} lemma,
saying that any renaming can be extended to the right by a context
\AgdaBound{$\Theta$}~\footnote{%
  We only require extension to the right because our syntax only has binding
  on the right.
  We could also extend on the left.
}.
To produce such an extended renaming, we receive a variable in \DTh{}, with the
two cases being that this variable is either in \AgdaBound{$\Delta$} or
\AgdaBound{$\Theta$}.
In the first case, we can use the original renaming \AgdaBound{$\rho$} to get
a variable in \AgdaBound{$\Gamma$}, which can be extended in the obvious way
to a variable in \GTh{}.
In the second case, we have a variable in \AgdaBound{$\Theta$}, and just need
to extend it to be a variable in \GTh{}.

\bindRen{}
\rename{}

We use renaming to show that, like variables, we can extend the context of terms
to the right.
The operation below renames a term by replacing all variables with variables
pointing to the left of the term's new context \GD{}.

@@

The action of a substitution on a term is given below.
The structure is very similar to that of renaming.
The differences are the following:

\begin{itemize}
  \item A substitution gives us terms, rather than variables.
        This means that when we use the substitution on a variable, we already
        have a term, and don't need to do any embedding into terms.
  \item When dealing with newly bound variables in \AgdaFunction{bindSub}, we
        need to produce terms, so we embed the newly bound variables into terms
        using \AgdaInductiveConstructor{var}.
  \item We use \AgdaFunction{$\swarrow^t$} instead of
        \AgdaFunction{$\swarrow^v$}, because we are dealing with terms rather
        than variables.
\end{itemize}

\bindSub{}
\substitute{}

\subsection{Kits}

To abstract over the similarities between renaming and substitution, we can use
\emph{kits} as introduced by McBride~\cite{McBride05,BHKM12}.

>>>>

@@

In the derivation below, we assume the existence of a derivation of
$B, C \to C \vdash C$, and by the admissibility of substitution we thus have a
derivation of $A \to B, A \vdash C$.
Intuitively, the context $A \to B, A$ derives the context $B, C \to C$, so
anything derived from $B, C \to C$ can also be derived from $A \to B, A$.
We see formally that $A \to B, A$ derives $B, C \to C$ by deriving each element
of the latter from the former --- hence the first two premises of the Subst
rule below, deriving $B$ and $C \to C$ from $A \to B, A$.

\begin{align*}
  &\begin{prooftree}
    \hypo{\Pi}
    \infer[no rule]1{A \to B, A \vdash B}
    \infer0[Var]{A \to B, A, C \vdash C}
    \infer1[$\to$-I]{A \to B, A \vdash C \to C}
    \hypo{B, C \to C \vdash C}
    \infer3[Subst]{A \to B, A \vdash C}
  \end{prooftree}
  \\\\
  &\textrm{where }\Pi \coloneqq
  \begin{prooftree}
    \infer0[Var]{A \to B, A \vdash A \to B}
    \infer0[Var]{A \to B, A \vdash A}
    \infer2[$\to$-E]{A \to B, A \vdash B}
  \end{prooftree}
\end{align*}

\subsection{Proofs of admissibility of renaming and substitution}

@@

We define \AgdaFunction{ren} as the action of a renaming on terms.
The idea of simultaneous renaming is to preserve the structure of the term, but
replace all of the variables from \AgdaBound{$\Delta$} by variables from
\AgdaBound{$\Gamma$}, with the mapping given by the renaming \AgdaBound{$\rho$}.

\rename{}

The \AgdaInductiveConstructor{var} case is where the action of the renaming
happens: the variable \AgdaBound{x} from \AgdaBound{$\Delta$} is mapped to the
variable \AgdaBound{$\rho$}\AgdaSpace{}\AgdaBound{x} from \AgdaBound{$\Gamma$}.
In the \AgdaInductiveConstructor{app} case, we have terms \AgdaBound{M}
\AgdaSymbol{:} \AgdaBound{$\Delta$} \AgdaDatatype{$\vdash$}
\AgdaBound{A} \AgdaInductiveConstructor{`$\to$} \AgdaBound{B} and \AgdaBound{N}
\AgdaSymbol{:} \AgdaBound{$\Delta$} \AgdaDatatype{$\vdash$} \AgdaBound{A}.
We may apply \AgdaFunction{ren} \AgdaBound{$\rho$} recursively to both
\AgdaBound{M} and \AgdaBound{N} to change their contexts from
\AgdaBound{$\Delta$} to \AgdaBound{$\Gamma$}, and the
\AgdaInductiveConstructor{app} constructor then produces the desired term in
\AgdaBound{$\Gamma$}.
Finally, in the \AgdaInductiveConstructor{lam} case, we get a term
\AgdaBound{M} \AgdaSymbol{:} \DA{} \AgdaDatatype{$\vdash$} \AgdaBound{B} and,
after introducing a \AgdaInductiveConstructor{lam} on the right, are in need
of a term of type \GA{} \AgdaDatatype{$\vdash$} \AgdaBound{B}.
To recursively apply \AgdaFunction{ren} to \AgdaBound{M}, we must thus extend
the renaming \AgdaBound{$\rho$} \AgdaSymbol{:} \RenGD{} with the newly bound
variable.
For this, we need an auxiliary function \AgdaFunction{bindRen} such that
\AgdaFunction{bindRen} \AgdaBound{$\rho$} \AgdaSymbol{:} \RenGADA{}.
This new renaming will act like \AgdaBound{$\rho$} for variables in
\AgdaBound{$\Delta$}, and map the new variable of type \AgdaBound{A} to the
corresponding new variable in \GA{}.

\bindRen{}

The \AgdaFunction{bindRen} given here has a slightly generalised type, where
instead of binding just a single variable of type \AgdaBound{A}, we could
bind a whole context \AgdaBound{$\Theta$} of new variables.
The first case of \AgdaFunction{bindRen} is for old variables from
\AgdaBound{$\Delta$}, where we apply \AgdaBound{$\rho$} to get a variable in
\AgdaBound{$\Gamma$}, and then use \AgdaFunction{$↙ᵛ$} to embed that variable
into \GTh{}.
The second case is for new variables from \AgdaBound{$\Theta$}, which embed
straight into \GTh{}.

The \AgdaFunction{sub} function below gives the action of a substitution.
Similarly to renaming, we want to preserve the structure of the term, except
now variables in the original term are replaced by \emph{terms} in the new
context.

\substitute{}

Given that this time, \AgdaBound{$\rho$} is a substitution rather than a
renaming, \AgdaBound{$\rho$} \AgdaBound{x} is a term, and is sufficient in the
\AgdaInductiveConstructor{var} case.
The \AgdaInductiveConstructor{app} case again deals with the subterms
recursively and then recombines them with \AgdaInductiveConstructor{app}.
In the \AgdaInductiveConstructor{lam} case, we again have a mismatch if we
want to apply \AgdaFunction{sub} recursively to the subterm \AgdaBound{M} with
an extra free variable.
We have \AgdaBound{$\rho$} \AgdaSymbol{:} \SubGD{} but need a substitution of
type \SubGADA{}, so we introduce the auxiliary definition
\AgdaFunction{bindSub}.

\bindSub{}

For the old variables in the first case, we have \AgdaBound{$\rho$} to turn
them into terms in \AgdaBound{$\Gamma$}.
Turning a term in \AgdaBound{$\Gamma$} into a term in \GTh{} requires a form
of weakening we have not yet proved, so I write \AgdaFunction{$↙ᵗ$} in analogy
with \AgdaFunction{$↙ᵛ$}, and prove it below.
In the second case, we want to substitute the new variable by the \emph{term}
referring to this new variable in \GTh{}.

The final piece to define substitution is to define the function that weakens
a term by some newly bound variables \AgdaBound{$\Delta$}.
For this, we use the action of renaming, which we have fully defined already,
and in particular rename each variable in the term from a variable in
\AgdaBound{$\Gamma$} to a variable in \GD{}.

@@

With this, the action of substitution is defined, and depends on the action
of renaming.

\subsection{Syntactic kits}\label{sec:syntactic-kits}

The statements of simultaneous renaming and simultaneous substitution are
very similar, with substitution being the generalisation that allows
replacement by terms rather than just variables.
I will introduce a type family \AgdaFunction{Env} of \emph{environments}, and
redefine \AgdaFunction{Ren} and \AgdaFunction{Sub} as environments of
variables and terms, respectively.

\Env{}
\RenSub{}

The processes I described for constructing proofs of the admissibility of
renaming and substitution were also similar.
Indeed, when we line up the resulting functions, \AgdaFunction{ren} and
\AgdaFunction{sub}, and their auxiliaries, \AgdaFunction{bindRen} and
\AgdaFunction{bindSub}, we notice only three key
differences:~\cite{McBride05,BHKM12}

\begin{itemize}
  \item In the first cases of \AgdaFunction{bindRen} and \AgdaFunction{bindSub},
    we do \AgdaFunction{$↙ᵛ$} and \AgdaFunction{$↙ᵗ$}, respectively, based on
    whether we are weakening a variable or a term.
  \item In the second case of \AgdaFunction{bindSub}, we do an extra wrapping of
    the new variable by \AgdaInductiveConstructor{var}, so as to make it a term
    to go in the substitution.
  \item In the \AgdaInductiveConstructor{var} case of \AgdaFunction{ren}, we
    have \AgdaInductiveConstructor{var} \AgdaSymbol(\AgdaBound{$\rho$}
    \AgdaBound{x}\AgdaSymbol) rather than just \AgdaBound{$\rho$} \AgdaBound{x},
    because the renaming \AgdaBound{$\rho$} gives us a variable rather than a
    term.
\end{itemize}

We may abstract over these three differences using the record \AgdaRecord{Kit}.
As in \AgdaFunction{Env}, we think of \AgdaBound{K} as being either
\AgdaDatatype{\_$\ni$\_} or \AgdaDatatype{\_$\vdash$\_}.
The fields of \AgdaRecord{Kit} are given in the same order as the points of
difference above.
Wherever the difference was presence or absence of
\AgdaInductiveConstructor{var}, we will be able to fill that field with either
\AgdaInductiveConstructor{var} or the identity function \AgdaFunction{id}.

\Kit{}

The field \AgdaField{$\swarrow^k$} can be seen as a property of the
judgement form \AgdaBound{K}, saying that it supports a form of weakening.
We use \AgdaField{vr} when adding a newly bound variable to an environment, and
use \AgdaField{tm} when we do a lookup from an environment and want to get a
term out.
Given a \AgdaRecord{Kit} \AgdaBound{K}, we can write the syntactic traversal
function \AgdaFunction{trav} and its auxiliary \AgdaFunction{bindEnv}, in the
model of \AgdaFunction{ren}, \AgdaFunction{sub}, and their auxiliaries.

\trav{}
\bindEnv{}

Concrete kits can be given for variables and terms either by inspecting
\AgdaFunction{ren} and \AgdaFunction{sub} or by following the types.
Notice that the kit for terms requires the admissibility of renaming so as to
achieve weakening of a substitution by newly bound variables.
Fortunately, this can be the \AgdaFunction{ren} defined below in terms of
\AgdaFunction{trav}, so we can keep \AgdaFunction{trav} as the only syntactic
traversal we have to write.

\begin{multicols}{2}
  \noindent\renKit{} \columnbreak

  \noindent\subKit{}
\end{multicols}
